---
title: false 
author: "dvarney - NOAA_StormAnalysis.Rmd"
date: "Friday, September 14, 2014"
output: html_document
---


####   $$Project 2: Reproducible Research$$
      
        
#### $Severe Weather And Its Impact On Public Health and Economy$ 

#### Abstract  
In this report, the aim is to analyze the impact of disparate and connected weather events on public health and the economy. Reliance is on the storm database of the U.S. National Oceanic and Atmospheric Administration's (NOAA) from 1950 to 2011. Estimates of fatalities, injuries, property and crop damage were used to decide which types of event are most harmful to the population health and economy. Excessive heat and tornados were found to be most harmful with respect to population health, while flood, drought, and hurricane/typhoon have the greatest economic consequences.

#### Introduction
Storms and other catastrophic weather events cause both public health and economic problems. for communities and municipalities. Many storm events cause person injuries and fatalities with property damage a secondary concern. Preventing such outcomes to the extent possible is a major concern.

This project involves exploring the NOAA storm database. This utility allows tracking characteristics of storms events, including time and location of estimates of fatalities, injuries, and property damage.
Data

Storm Data is available as a downloadable file of 47Mb (megabytes). A document pairs with the data to provide required knowledge about the use and partitioning of the data file. For assistance, a Frequently Asked Qustions (FAQ) web page is available.

Database events initiated approximately 1950, ending in November 2011. Fewer events were recorded for years prior to 1995, which is most likely due to a lack of diligent record-keeping. Dates after this time period were considered as more complete and used for the report.

The methods section encompases the data analysis and addresses two, primary questions. 1.) Regarding the United States, which types of events are most harmful with respect to population health, and 2.) which of events have the greatest economic consequences?

To assist in answering these issues, figures are included to illustrate points in the report. Recommendations are not suggested, but the presentation of statistically inferred data analysis is significant. For transparency, the data source is available from NOAA and R-code can be acquired in its entirety from the author.  

##### Initial settings and load libraries
```{r, message=FALSE}
library(R.utils)
library(ggplot2)
library(plyr)
require(gridExtra)

warning = FALSE
echo = TRUE           # Globally, make code visible
options(scipen = 2)   # No scientific notation, but 2 dp
```   


#### Data Processing

Redirect R-code and data file store.
If the data already exists, we do not need to load it again. Otherwise, download and unzip the file.
```{r}
setwd("K:/COURSES/JHU_DataScience/ReproducibleResearch/Project_2")

#is there a stormData file, if not then download at NOAA's website and unzip
#also, display a progress bar
if (!"NOAA_StormData.csv.bz2" %in% dir("data"))
{
  message("PLZ WAIT, downloading ... this may take several minutes")
  download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", 
                destfile="data/NOAA_StormData.csv.bz2", 
                quiet=FALSE)
}

#unzip data file to 'NOAA_StormData.csv'
if (!"NOAA_StormData.csv" %in% dir("data"))
{
    message("unzipping data file")
    bunzip2("NOAA_StormData.csv.bz2", overwrite=FALSE, remove=FALSE)
}

```


check for a 'stormData' datafile loaded, then read if available
```{r, warning=FALSE}
if (!"stormData" %in% ls()) 
{
    message("reading NOAA data file")
    stormData <- read.csv("data/NOAA_StormData.csv", sep = ",", stringsAsFactors=FALSE)
}

message("READY to process")
```
Inspect a small section of the 'RAW' data file, there are 37 possible columns of information.

```{r}
head(stormData, n = 2)
```
The events in this database start in April 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, possibly due to lacks records keeping. Recent years have better coverage.
A quick scan of the data suggests that the number of storm events did increase during this period. A subset from 1995-2011 was extracted for this analysis.

```{r}

  stormData$year <- as.numeric(format(as.Date(stormData$BGN_DATE, 
                                                  format = "%m/%d/%Y %H:%M:%S"), "%Y"))

  #hist(stormData$year, breaks=30, col="lightblue", main="Storms by Year", xlab="Year")
  storm <- stormData[stormData$year >= 1995,]

```  
Remove the above comment control to view the number of storms over the period 1950-2011

##### Methods
In this section, we check the number of personal **fatalities** and **injuries** that were caused by severe weather. The top 12 of the most severe types of weather events are reviewed
```{r}
fSortAssist <- function(fieldName, top=12, data=stormData) 
  {
    index <- which(colnames(data) == fieldName)
    field <- aggregate(data[, index], by=list(data$EVTYPE), FUN="sum")
    names(field) <- c("EVTYPE", fieldName)
    field <- arrange(field, field[, 2], decreasing=TRUE)
    field <- head(field, n=top)
    field <- within(field, EVTYPE <- factor(x=EVTYPE, levels=field$EVTYPE))
    return(field)
  }

message("Stand By, processing data ...")
fatalities <- fSortAssist("FATALITIES", 12, data=storm)
injuries <- fSortAssist("INJURIES", 12, data=storm)

```

###### Impact on Economy
I convert the **property** and **crop damage** data into a monetary form according to the units described in the NOAA document, ([NWSPD 10-16, NATIONAL WEATHER SERVICE INSTRUCTION  10-1605](, available at: http://www.nws.noaa.gov/directives/)). Both *PROPDMGEXP* (PROPerty DaMaGe EXPense) and *CROPDMGEXP* (CROP DaMaGe EXPense) columns record a cost multiplier for each event, where Hundred (H), Thousand (K), Million (M) and Billion (B) have their respective meaning.

```{r, warning=FALSE}
#cost x Hundred (H), Thousand (K), Million (M) and Billion (B).
fMonetaryAssist <- function(data = storm, fieldName, newFieldName) 
  {
    t_Len <- dim(data)[2]
    index <- which(colnames(data) == fieldName)
    data[, index] <- as.character(data[, index])
    qNA <- !is.na(toupper(data[, index]))
    
    data[qNA & toupper(data[, index]) == "B", index] <- "9"   #billions
    data[qNA & toupper(data[, index]) == "M", index] <- "6"   #millions
    data[qNA & toupper(data[, index]) == "K", index] <- "3"   #thousands
    data[qNA & toupper(data[, index]) == "H", index] <- "2"   #hundreds
    data[qNA & toupper(data[, index]) == "" , index] <- "0"
    
    data[, index] <- as.numeric(data[, index])
    data[is.na(data[, index]), index] <- 0
    data <- cbind(data, data[, index - 1] * 10^data[, index])
    
    
    names(data)[t_Len + 1] <- newFieldName
    return(data)
  }

    storm <- fMonetaryAssist(storm, "PROPDMGEXP", "propertyDamage")
    storm <- fMonetaryAssist(storm, "CROPDMGEXP", "cropDamage")
    names(storm)
    options(scipen=999)

    property <- NULL
    crop <- NULL
    property <- fSortAssist("propertyDamage", 12,  data=storm)
    crop <- fSortAssist("cropDamage", 12, data=storm)
scan()
```

#### Results
For the impact on public health, we have two lists of severe weather events, listed below. For the 12 most severe, the number of people affected are enumerated. Of note are the numbers for **Excessive Heat** and **Tornados**.

Here, the two tables, FATALITIES and INJURIES, are merged. This allows viewing the physical (bodily) 'damage' side-by-side, assisting in comparision for probable events outcomes.

```{r}
merge(fatalities, injuries)

```
Below is a set of graphs of fatalities and injuries from severe weather events. 
```{r}
#fatalities for 12 severe weather events summed across 1995-2011
pFatalities <- qplot(EVTYPE, data=fatalities, weight=FATALITIES, binwidth = 1) +
                    theme_bw() + 
                    geom_bar(aes(fill="red"), color="black") +
                    scale_y_continuous("Fatalities") + 
                    theme(axis.text.x = element_text(angle = 45, hjust = 1), 
                          legend.position = "none") +
                    xlab("")
 
#injuries, with the same parameters
pInjuries <- qplot(EVTYPE, data=injuries, weight=INJURIES, binwidth = 1) + 
                  theme_bw() + 
                  geom_bar(fill="lightblue", color="black") +
                  scale_y_continuous("Injuries") + 
                  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
                        legend.position = "none") +
                   xlab("")

#create a panel plot of the above two
grid.arrange(pFatalities, pInjuries, ncol = 2,
             main="Severe Weather Events, 1995-2011\nFatalities and Injuries")
```

Find all weather events that are either hail or tornado related. As there is a strong correlation between these factors, the data is scoured by state over the 1995-2011 time frame.

```{r} 
evt_dim <- 1:55
evt_names <- c("count1", "count2", "where", "StNum", "StName")
evt <- data.frame(evt_dim, evt_dim, evt_dim, evt_dim, evt_dim)
colnames(evt) <- evt_names
evt$StName <- NA


for (i in 1:55)
  {
    evt$count1[i] <- length(which(storm[,"EVTYPE"] == "HAIL" & storm[,"STATE__"] == i))
    evt$count2[i] <- length(which(storm[,"EVTYPE"] == "TORNADO" & storm[,"STATE__"] == i)) 
    evt$where[i]  <- i
  }

for (i in 1:nrow(storm))
  {
    idx <- storm[i,"STATE__"]
    if (is.na(evt$StName[idx]))
      {
        evt$StName[idx] <- as.character(storm[i, "STATE"])
        evt$StNum[idx]  <- i
      }
    if (idx == 55) break
  }

```
Scan the data for non-weather events, SUMMARY and Metro and remove
```{r}
unq <- unique(storm$EVTYPE)
unq[grep("^Summ|^SUMM|^Metr", unq)] = NA   # remove all 'summary|SUMMARY|Metro', keep others
unq <- unq[!is.na(unq[])] 
head(unq, 20)
```
Find all the Hail and Tornado events
```{r}
Hail_Events <- unq[grep("^HAIL", unique(storm$EVTYPE))]
Tornado_Events <- unq[grep("^TORN", unique(storm$EVTYPE))]
```
Show a few types of Hail and Tornado events
```{r}
Hail_Events
Tornado_Events
```

This plot depicts the total hail and tornado events across the country during the study period. The plot is a bit complicated, though a shrewd eye may deciper the details.

```{r}
torn = "red"
hail = "lightblue"

t_HailTornadoByState<-ggplot() + theme_bw() + 
                  geom_bar(data=evt, aes(x=StName, y=count2/100, color=torn), 
                           fill=torn, stat="identity", alpha=0.5) + #tornado
                  geom_bar(data=evt, aes(x=StName, y=count1/1000, color=hail), 
                           fill=hail, stat="identity", alpha=0.8) + #hail
                  scale_y_continuous("HAIL / TORNADO Events\n(x1000, x100)") +                
                  theme(axis.text.x = element_text(angle = 90, hjust = 1), 
                        legend.position=c(0.1, 0.825)) +
                  scale_color_manual(name = "EVENT", values = c("black", "green"), labels=c("HAIL", "TORNADO")) +
                 
                  theme(legend.background = element_rect(color = 'purple', fill = 'white', size = 1)) +
                  scale_fill_manual(values=c(hail="red", torn="blue"), 
                                    guide="none") +
                  xlab("State") 

print(t_HailTornadoByState)
```  

Based on the above histograms, we find that **excessive heat** and **tornados** cause most fatalities; tornados caused most injuries in the United States from 1995 to 2011. This seems counter intuitive, as one might consider hurricanes the most damaging and dangerous. However, tornados appearances can have very little warning, especially during the night, when most individuals are not quite awake when they hear a siren or the roar of the high velocity winds. 

```{r}


printMoney <- function(x)
  {
    format(x, digits=10, nsmall=2, decimal.mark=".", big.mark=",")
  }
                   
t_property <- property
t_crop<- crop

#change column names to '$'name
npd <- names(t_property)[2]; new_npd <- paste("$", npd, sep=""); names(t_property)[2] <- new_npd
ncd <- names(t_crop)[2];     new_ncd <- paste("$", ncd, sep=""); names(t_crop)[2] <- new_ncd

#select weather events common to property and crop damage
mpc <- merge(t_property, t_crop) 

#calc ratio between property and crops
mpc[,4] <- round(mpc[,2]/mpc[,3], 2)
colnames(mpc)[4] <- "ratio"

#format property and crop numbers to $$ amounts with '.,'
mpc[,2] <- printMoney(mpc[,2])
mpc[,3] <- printMoney(mpc[,3])
```
For the economic impact, see the table below. The list displays the common damage events, e.g. FLOOD or   between the property and crops. There is a large difference in damage estimates, as noted by the 'ratio' column in the data table of monetary damages.

```{r}
mpc

```
The following is a pair of graphs of total property damage and crop damage affected by the mentioned severe weather events. 
```{r}
propertyDamage <- qplot(EVTYPE, data=property, weight=propertyDamage/1e+09, binwidth=1) + 
                        theme_bw() + theme(legend.position = "none") +
                        geom_bar(aes(fill="lightblue", color="black")) +    
                        theme(axis.text.x=element_text(angle=45, hjust=1)) + 
                        ylab(expression(paste("Property Damage in US dollars (1x ",10^9,")", sep=""))) +
                        xlab("")

cropDamage <- qplot(EVTYPE, data=crop, weight=cropDamage/1e+09, binwidth=1) + 
                    theme_bw() + theme(legend.position = "none") +
                    geom_bar(aes(fill="lightblue"), color="black") +
                    theme(axis.text.x=element_text(angle=45, hjust=1)) +
                    ylab(expression(paste("Crop Damage in US dollars (1x ",10^9,")", sep=""))) +
                    xlab("")
             
grid.arrange(propertyDamage, cropDamage, ncol = 2,
             main=("Severe Weather Events, 1995-2011\nProperty and Crop Damage in the United States (Billions)"))
```  

Based on the above histograms, we find that **flood** and **hurricane/typhoon** cause most property damage; **drought** and **flood** causes most crop damage in the United States from 1995 to 2011.

#### Conclusion  
From these data, we found that **excessive heat** and **tornado** are most harmful with respect to population health, while **flood**, **drought**, and **hurricane/typhoon** have the greatest economic consequences.
